Ultrafast

### 2023-08-06 manuscript/manuscript.qmd
Process Robert comments

---
author: 
  - name: "Kjell Jorner"
    orcid: 0000-0002-4191-6790
    email: "kjell.jorner@chem.ethz.ch"
    corresponding: true
    affiliations: 
      - ref: eth
      - ref: chalmers
      - ref: toronto-chem
      - ref: toronto-cs
  - name: "Robert Pollice"
    orcid: 0000-0001-8836-6266
    email: "r.pollice@rug.nl"
    corresponding: true
    affiliations: 
      - ref: rug
      - ref: toronto-chem
      - ref: toronto-cs  
  - name: "Cyrille Lavigne"
    orcid: 0000-0003-2778-1866
    affiliations: 
      - ref: toronto-chem
      - ref: toronto-cs    
  - name: "Alán Aspuru Guzik"
    orcid: 0000-0002-8277-4434
    affiliations:
      - ref: toronto-chem
      - ref: toronto-cs
      - ref: toronto-chem-eng
      - ref: toronto-mat-eng
      - ref: vector
      - ref: ac

@@

Organic light-emitting diodes (OLEDs) is a technology for generating light from electricity using organic molecules.[@yangRecentAdvancesOrganic2017] The first generation of OLEDs were limited to an efficiency of max 25% due to spin statistics. As the transition T~1~ $\rightarrow$ S~0~, from the lowest excited triplet state to the singlet ground state, is quantum-mechanically forbidden, the OLED molecule can formally only emit from its first excited state of singlet multiplicity, S~1~. The second generation of OLEDs exploited ways of increasing the rate of this spin-forbidden phosphorescence. The third generation of OLEDs are based on Thermally Activated Delayed Fluorescence (TADF), where the S~1~ state is partially populated from a near-lying T~1~ state. For this to happen, the gap $\DeltaE{ST}$ between the states needs to sufficiently small so that the thermal activation competes with non-radiative decay processes from T~1~. However, as the T~1~ population is still substantial, problems with stability and lower than ideal intensity due to non-radiative decay persist. The logical step for the next generation of OLEDs is to emit directly from an S~1~ state that lies below the T~1~ state, potentially converting all of the electrically generated excitons into photons, _i.e._, 100% efficiency.

Molecules that fulfill this criterion have been branded with the acronym INVErted Singlet Triplet energy gap (INVEST) and are exceedingly rare.[@liOrganicMoleculesInverted2022] They are said to violate Hund's rule of maximum multiplicity, which describes that the electronic state with the highest spin (in this case the triplet) should be lowest in energy. While examples of molecules violating Hund's rule in the excited state have been known since the 1970s and 1980s,[@kollmarViolationHundRule1978; @leupinLowlyingElectronicallyExcited1980; @kosekiViolationHundMultiplicity1985] it is only recently that they have received considerable attention for application in OLEDs.[@desilvaInvertedSingletTriplet2019;@polliceOrganicMoleculesInverted2021;@aizawaDelayedFluorescenceInverted2022] As they are rare, recent efforts have used high-throughput virtual screening (HTVS) with computational chemistry to identify new compounds with inverted gaps, focusing on expanding the local chemical space around specific scaffolds[@polliceOrganicMoleculesInverted2021;@aizawaDelayedFluorescenceInverted2022] or scanning larger (combinatorially or experimentally derived) datasets to identify novel scaffolds.[@terenceblaskovitsSymmetryInducedSinglet2023;@omarIdentificationUnknownInverted2023;@garnerDoublebondDelocalizationNonalternant2023] These HTVS efforts have been successful at identifying several new INVEST candidates, some of which have also been synthesized and tested experimentally.[@aizawaDelayedFluorescenceInverted2022]

@@

Motivated by the need for faster methods for HTVS of INVEST compounds, we wondered if it would be possible to perform much simpler calculations as a pre-screening step for the more expensive methods. Based on prior work in the literature from the 1970s and 1980s, we find that the inverted gaps can be recovered already with the semi-empirical Pariser-Parr-Pople method using configuration interaction singles (CIS) and adding key double excitations. The PPP method considers only the $\pi$ electrons in a minimal valence basis and approximates all integrals from experimental data and a few empirical parameters, making it computationally extremely cheap. At the same time, it retains the conceptual clarity of Hückel theory and allows for straightforward interpretation of the inverted gap in terms of the well-known concept of dynamic spin polarization (DSP). We show that the method is capable of finding hits both in the local chemical space around known scaffolds, and of identifying hits in more diverse datasets. While there are clear limitations to the method, we believe that it will open the doors for ultra-high-throughput virtual screening (UHTVS) of compound libraries orders of magnitude larger than today. The method is also perfectly applicable as a scoring function in inverse design algorithms.

@@

While the recent literature has focused on applying high-level *ab initio* and DFT methods to study INVEST compounds, we believe that some conceptual clarity has been lost in process. Instead, we apply the simplest possible electronic structure method that can capture the physics of the problem. The Pariser-Parr-Pople (PPP) method is an extension of the semi-empirical Hückel molecular orbital theory for $\pi$-electron systems that includes electron correlation.[@pariserSemiEmpiricalTheory1953a;@pariserSemiEmpiricalTheory1953; @popleElectronInteractionUnsaturated1953] It originated in the 1950s, with new parametrizations being developed mainly in the 1960s and 1970s, and was used in the dye industry at least until the 1980s before the advent of more accurate ab initio methods.[@griffithsPracticalAspectsColour1982] The PPP method retains the conceptual clarity of Hückel theory, while at the same time including the electron correlation that is necessary to capture the inverted gap.

@@

where $\alpha_{r}$ is the *one-center core resonance integral*, $U_r$ is a parameter of model called the *atomic valence state potential*, $Z_s$ is the effective nuclear charge and $\beta_{rs}$ is the *two-center resonance integral*, another parameter of the model.

@@

where $K$ is the exchange integral between the two orbitals involved in the single excitation (normally the HOMO and the LUMO). As defined here, a positive $\DeltaE{ST}$ corresponds to the usual situation with the triplet being lower in energy than the singlet, while a positive $\DeltaE{ST}$ corresponds to an inverted gap. The exchange integral depends strongly on the spatial overlap between the the two orbitals, which in the ZDO approximation can be computed as:[@peachExcitationEnergiesDensity2008]

@@

where $|c_{r,i}|$ is the absolute value of the coefficient of orbital $i$ on center $r$. When the overlap is zero, the exchange interaction vanishes and $\DeltaEST{SCF}$ is zero. Unfortunately, so is the oscillator strength, $f$, between the two states,[@klessingerExcitedStatesPhotochemistry1995] which is non-optimal for OLED materials that should emit light. Consequently, there is a trade-off between having a small HOMO-LUMO overlap to reduce the exchange interaction, while still maintaining a sufficient oscillator strength.

@@

We used four separate datasets for this study. The first is a set of azaphenalenes previously studied by some of us, comprising 256 substituted compounds, for which excitation energies were computed with (2nd-order) algebraic-diagrammatic construction, {{< var reference_level >}}/[Robert: Fill out], and geometries were optimized with B3LYP/[Robert: Fill out][@polliceOrganicMoleculesInverted2021] The second is a set substituted 138 azaazulenes with {{< var reference_level >}}/[Robert: Fill out] excitation energies and optimized with the B97-3c composite method.[Robert: Add ref to dataset if it exists.] The third is a set of 16 rationally designed scaffolds, which including substitutions amounts to {{< var global_n_total >}} unique compounds, optimized at the B97-3c level and with excitation energies at the {{< var reference_level >}}/[Robert: Fill out] level.[Robert: Add ref to dataset if it exists.] The fourth is a set 315 non-alternant hydrocarbons including substitutions, divided into three subsets of size 76, 187 and 52, optimized at the $\omega$B97X-D/def2-TZVP level and with excitation energies at the CC2/aug-cc-pVTZ level.

::: {.callout-important}
Robert: I have 138 azazulenes with data in the csv file but 144 xyz files.
:::

@@

## Qualitative understanding of the inverted gap

@@

![Orbitals energies and frontier orbitals of pentalene at the idealized $D_\mathrm{2h}$ geometry and bond lengths of 1.4 Å.]({{< var fig_pentalene_ideal_orbitals >}}.png){#fig-pentalene-ideal-orbitals}

@@

![Orbitals energies and frontier orbitals of pentalene at the optimized $C_\mathrm{2h}$ geometry.]({{< var fig_pentalene_orbitals >}}.png){#fig-pentalene-orbitals}

@@

Azaphenalene is the prototypical molecule that started the renewed investigations into INVEST molecules,[@desilvaInvertedSingletTriplet2019;@sobolewskiAreHeptazineBasedOrganic2021] and has been the subject of numerous studies with high-level quantum chemical methods. Can we capture the inverted gap, as first measured by Leupin and Wirz in 1980?[@leupinLowlyingElectronicallyExcited1980] The calculations reveal that the HOMO and LUMO are well separated with spatial overlap of {{< var azaphenalene_overlap >}} and a correspondingly low exchange interaction of only {{< var azaphenalene_2_exchange >}} kcal/mol ([@fig-azaphenalene-orbitals]). The larger azaphenalene has 30 excitations that could possibly contribute to the large $\DeltaE[DSP]{SCF}$ of {{< var azaphenalene_dsp_scf >}} kcal/mol, but the most import by far are those from the doubly degenerate HOMO-1 to the doubly degenerate LUMO+1 ([@fig-azaphenalene-excitations]). In total, $\DeltaEST[DSP]{SCF}=$ {{< var azaphenalene_t1_s1_dsp_scf >}} kcal/mol, which is slightly smaller at the CIS level with $\DeltaEST[DSP]{CIS}=$ {{< var azaphenalene_t1_s1_dsp_cis >}} kcal/mol.

To conclude this section on the model compounds, our PPP-based protocol is able to capture the inverted gap, and also provides qualitative understanding into the physical mechanism of dynamic spin polarization through identification and visualization of the corresponding excitations. For these two compounds, the calculated $\DeltaEST[DSP]{}$ are more negative at the SCF level compared to CIS. Compared to {{< var reference_level >}}, it would seem that CIS is preferable, but to elucidate to which method is better, we now turn to larger datasets.

![Orbitals energies and frontier orbitals of azaphenalene.]({{< var fig_azaphenalene_orbitals >}}.png){#fig-azaphenalene-orbitals}

@@

While our method captures the inverted gap of azaphenalene and pentalene, it also needs to capture trends with substitution to work effectively in virtual screening. We therefore calculated 256 substituted azaphenalenes that have previously been studied by some of us.[@polliceOrganicMoleculesInverted2021] The PPP level S~1~ and T~1~ excitation energies are well correlated with those from {{< var reference_level >}} with $R^2$ of {{< var local_azaphenalenes_r2_singlet >}} and {{< var local_azaphenalenes_r2_triplet >}}, respectively ([@fig-local-azaphenalenes]a,b). Unfortunately, the oscillator strengths are not as well reproduced with an $R^2$ of {{< var local_azaphenalenes_r2_f >}}, although the Spearman $\rho$ of {{< var local_azaphenalenes_rho_f >}} indicates the values might be used to rank potential candidates ([@fig-local-azaphenalenes]c). The problems of obtaining accurate oscillator strengths at the PPP level of theory are well known in the literature, and the problem is especially exacerbated with low oscillator strengths as here.[@misurkinElectronicStructureConjugated1974] Crucially, the $\DeltaE{ST}$ shows good correlations, with $R^2=$ {{< var local_azaphenalenes_r2_cis >}} using $\DeltaEST[DSP]{CIS}$ ([@fig-local-azaphenalenes]f). This $R^2$ value is essentially unchanged from the gap at the CIS level without DSP ({{< var local_azaphenalenes_r2_cis_raw >}}, [@fig-local-azaphenalenes]d) and markedly better than for the gap at the SCF level ({{< var local_azaphenalenes_r2_scf >}}, [@fig-local-azaphenalenes]e), showing the importance of going beyond the SCF level to include at least some configuration interaction. The results are of similar or better quality recently achieved by Pernal and co-workers for a different set of azaphenalenes using SCF + DSP with orbitals from DFT.[@drwalRoleSpinPolarization2023] To further analyze the results in terms of a binary classification into normal/inverted, we calculated the true positives (TP), true negatives (TN), false positives (FP) and false negatives (FN), as well as a range of common classification scores (@tbl-local-azaphenalenes). For definitions, see [@eq-f1] to [@eq-roc-auc]. The most important metrics for virtual screening are, in our opinion, the recall and the specificity. The recall measures the fraction of inverted molecules that the protocol captures, and is {{< var local_azaphenalenes_recall_scf >}} at the SCF level and {{< var local_azaphenalenes_recall_cis >}} at the CIS level. The specificity measures the fraction of non-inverted molecules that the protocol identifies, and is {{< var local_azaphenalenes_specificity_scf >}} at the SCF level and {{< var local_azaphenalenes_specificity_cis >}} at the CIS level. These metrics mean that we are able to capture a large proportion of the molecules with inverted gaps, while filtering out most non-inverted molecules. The results for these azaphenalenes indicate that the PPP protocol could be used to prescreen candidates of this compound class for $\DeltaE{ST}$, while further pruning with another method is likely needed for the oscillator strengths.

![Excitation properties for azaphenalenes _vs._ the reference level. (a) Singlet excitation energies. (b) Triplet excitation energies. (c) Oscillator strengths. (d) Singlet-triplet energy gaps with CIS. (e) Singlet-triplet energy gaps with SCF + DSP. (f) Singlet-triplet energy gaps with CIS + DSP.]({{< var fig_local_azaphenalenes >}}.pdf){#fig-local-azaphenalenes}

@@

While we see success for the azaphenalenes, the situation for the azaazulenes is unfortunately worse. The PPP level S~1~ and T~1~ excitation energies are only moderately correlated with those from {{< var reference_level >}} with $R^2$ of {{< var local_azulenes_r2_singlet >}} and {{< var local_azulenes_r2_triplet >}}, respectively ([@fig-local-azulenes]a,b). For the oscillator strengths, the situation is rather catastrophic with $R^2$ of {{< var local_azulenes_r2_f >}} implying no correlation whatsoever ([@fig-local-azulenes]c). The situation for the $\DeltaE{ST}$ is also worse than for the azaphenalenes, with $R^2=$ {{< var local_azulenes_r2_cis >}} with CIS + DSP, a marked improvement on {{< var local_azulenes_r2_scf >}} with SCF + DSP ([@fig-local-azulenes]e,f). Unfortunately, the negative gaps are not recovered, leading to recalls of {{< var local_azulenes_recall_cis >}} as none of five inverted molecules could be identified ([@tbl-local-azulenes]). It could be speculated that the worse performance for the azaazulenes comes from the fact that they are non-alternants, for which the approximations are expected to be less applicable. Considering the contrasting performance for the azaphenalenes and azaazulenes, it seems clear that the PPP screening protocol will inevitably miss some hits belonging to particular compound classes. To investigate the global performance of the model we turned to a more diverse set with 16 different compound classes.

![Excitation properties for azaazulenes _vs._ the reference level. (a) Singlet excitation energies. (b) Triplet excitation energies. (c) Oscillator strengths. (d) Singlet-triplet energy gaps with CIS. (e) Singlet-triplet energy gaps with SCF + DSP. (f) Singlet-triplet energy gaps with CIS + DSP.]({{< var fig_local_azulenes >}}.pdf){#fig-local-azulenes}

@@

Here we used a dataset previously generated by some of us using rational design rules, comprising 16 different molecular scaffolds, the details of which will be presented elsewhere. Out of {{< var global_n_total >}} compounds, only {{< var global_n_failed >}} failed to compute, which corresponds to a success rate of {{< var global_percent_success >}}%. To put this into context, ~25% of the CASSCF calculations and 9% of the CIS calculations failed in a recent HTVS workflow by Troisi, Padula and co-workers.[@omarIdentificationUnknownInverted2023] The total runtime using four cores on the M2 processor was {{< var time_cpu_hours >}} CPU hours, with the mean time per compounds at {{< var time_mean >}} seconds. The S~1~ and T~1~ excitation energies are fairly reproduced with ([@fig-global-all]a,b) $R^2$ of {{< var global_all_r2_singlet >}} and {{< var global_all_r2_triplet >}} vs {{< var reference_level >}}, respectively. As seen in [@fig-global-all-gaps]a, the $\DeltaE{ST}$ at the SCF + DSP level shows a moderate $R^2$ of {{< var global_all_r2_scf >}}, which increases significantly to {{< var global_all_r2_cis >}} at the CIS + DSP level ([@fig-global-all-gaps]b). However, due to a systematic overestimation of $\DeltaE{ST}$, the F1 score is still low at {{< var global_all_f1_cis >}} and the recall is only {{< var global_all_recall_cis >}} ([@tbl-global-all]). We therefore added a linear correction $\DeltaEST[DSP]{CIS,LC}=$ {{< var global_lc_coef >}} $\times\DeltaEST[DSP]{CIS}$ {{< var global_lc_intercept >}}, which increases the F1 score to {{< var global_all_f1_cis_corr >}} and the recall to {{< var global_all_recall_cis_corr >}} ([@fig-global-all-gaps]c). The specificity simultaneously decreases from {{< var global_all_specificity_cis >}} to {{< var global_all_specificity_cis_corr >}}, but overall the linear correction would be preferable for virtual screening where there is a strong focus on finding rare hits. Applying the linear correction to the azaphenalenes ([@fig-local-corr]a) and azaazulenes ([@fig-local-corr]b) leads to an improvement in the recall of the former from {{< var local_azaphenalenes_recall_cis >}} to {{< var local_azaphenalenes_recall_cis_corr >}} while still none of the inverted azaazulenes are recovered. The oscillator strengths are unfortunately poorly correlated for the rational design set, with $R^2=$ {{< var global_all_r2_oscillator >}}, although the Spearman rank correlation is more encouraging at {{< var global_all_rho_oscillator >}} ([@fig-global-all]c). Taken together, the results reinforce the conclusions from the case study of the azaphenalenes, that the PPP protocol is suitable for screening $\DeltaE{ST}$ while it struggles for oscillator strengths.

>>>>

---
author: 
  - name: "Kjell Jorner"
    orcid: 0000-0002-4191-6790
    email: "kjell.jorner@chem.ethz.ch"
    corresponding: true
    affiliations: 
      - ref: eth
      - ref: chalmers
      - ref: toronto-chem
      - ref: toronto-cs
  - name: "Robert Pollice"
    orcid: 0000-0001-8836-6266
    email: "r.pollice@rug.nl"
    corresponding: true
    affiliations: 
      - ref: rug
      - ref: toronto-chem
      - ref: toronto-cs  
  - name: "Cyrille Lavigne"
    orcid: 0000-0003-2778-1866
    affiliations: 
      - ref: toronto-chem
      - ref: toronto-cs    
  - name: "Alán Aspuru Guzik"
    orcid: 0000-0002-8277-4434
    affiliations:
      - ref: toronto-chem
      - ref: toronto-cs
      - ref: toronto-chem-eng
      - ref: toronto-mat-eng
      - ref: vector
      - ref: ac
      - ref: cifar

@@

Organic light-emitting diodes (OLEDs) is a technology for generating light from electricity using organic molecules.[@yangRecentAdvancesOrganic2017] The first generation of OLEDs were based on fluorescent organic molecules and limited to an efficiency of 25% due to spin statistics. As the transition T~1~ $\rightarrow$ S~0~, from the lowest excited triplet state to the singlet ground state, is spin-forbidden, the OLED molecule can only emit efficiently from its first excited state of singlet multiplicity, S~1~. The second generation of OLEDs exploited ways of increasing the rate of this spin-forbidden phosphorescence. The third generation of OLEDs are based on Thermally Activated Delayed Fluorescence (TADF), where the S~1~ state is partially populated from a near-lying T~1~ state. For this to happen, the gap $\DeltaE{ST}$ between the states needs to be sufficiently small so that the thermal activation competes with non-radiative decay processes from T~1~. However, as the T~1~ population is still substantial, problems with stability and lower than ideal quantum yields due to non-radiative decay persist. The logical step for the next generation of OLEDs is to emit directly from an S~1~ state that lies below the T~1~ state, potentially converting all of the electrically generated excitons into photons, _i.e._, 100% efficiency, while avoiding decay reactions from the T~1~ state.

Molecules that fulfill this criterion have been branded with the acronym INVErted Singlet Triplet energy gap (INVEST) and are exceedingly rare.[@liOrganicMoleculesInverted2022] They are said to violate Hund's rule of maximum multiplicity as applied to the S~1~ and T~1~ states, which describes that the electronic state with the highest spin (in this case the triplet) should be lowest in energy. While examples of molecules violating Hund's rule in the excited state have been known since the 1970s and 1980s,[@kollmarViolationHundRule1978; @leupinLowlyingElectronicallyExcited1980; @kosekiViolationHundMultiplicity1985;@toyotaViolationHundRule1988;@toyotaViolationHundMultiplicity1986] it is only recently that they have received considerable attention for application in OLEDs.[@desilvaInvertedSingletTriplet2019;@polliceOrganicMoleculesInverted2021;@aizawaDelayedFluorescenceInverted2022] As they are rare, recent efforts have used high-throughput virtual screening (HTVS) with computational chemistry to identify new compounds with inverted gaps, focusing on expanding the local chemical space around specific scaffolds[@polliceOrganicMoleculesInverted2021;@sanz-rodrigoNegativeSingletTriplet2021;@ricciSingletTripletExcited2021;@sobolewskiAreHeptazineBasedOrganic2021;@aizawaDelayedFluorescenceInverted2022;@ricciEstablishingDesignStrategies2022;@tuckovaOriginInvertedSinglet2022] or scanning larger (combinatorially or experimentally derived) datasets to identify novel scaffolds.[@terenceblaskovitsSymmetryInducedSinglet2023;@omarIdentificationUnknownInverted2023;@garnerDoublebondDelocalizationNonalternant2023] These HTVS efforts have been successful at identifying several new INVEST candidates, some of which have also been synthesized and tested experimentally.[@aizawaDelayedFluorescenceInverted2022]

@@

Motivated by the need for faster methods for HTVS of INVEST compounds, we wondered if it would be possible to perform much simpler calculations as a pre-screening step for the more expensive methods. Based on prior work in the literature from the 1970s and 1980s, we find that the inverted gaps can be recovered already with the semi-empirical Pariser-Parr-Pople method using configuration interaction singles (CIS) and adding key double excitations. The PPP method considers only the $\pi$ electrons in a minimal valence basis and approximates all integrals from experimental data and a few empirical parameters, making it computationally extremely cheap. At the same time, it retains the conceptual clarity of Hückel theory and allows for straightforward interpretation of the inverted gap in terms of the well-known concept of dynamic spin polarization (DSP). We show that the method is capable of finding hits both in the local chemical space around known scaffolds, and of identifying hits in more diverse datasets. While there are clear limitations to the method, we believe that it will open the doors for ultra-high-throughput virtual screening (UHTVS) of compound libraries several orders of magnitude larger than today. The method is also perfectly applicable as a scoring function in inverse design algorithms.

@@

While the recent literature has focused on applying high-level *ab initio* and DFT methods to study INVEST compounds, we believe that some conceptual clarity has been lost in the process. Instead, we apply the simplest possible electronic structure method that can capture the physics of the problem. The Pariser-Parr-Pople (PPP) method is an extension of the semi-empirical Hückel molecular orbital theory for $\pi$-electron systems that includes electron correlation.[@pariserSemiEmpiricalTheory1953a;@pariserSemiEmpiricalTheory1953; @popleElectronInteractionUnsaturated1953] It originated in the 1950s, with new parametrizations being developed mainly in the 1960s and 1970s, and was used in the dye industry at least until the 1980s before the advent of more accurate ab initio methods.[@griffithsPracticalAspectsColour1982] The PPP method retains the conceptual clarity of Hückel theory, while at the same time including the electron correlation that is necessary to capture the inverted gap.

@@

where $\alpha_{r}$ is the *one-center core resonance integral*, $U_r$ is a parameter of the model called the *atomic valence state potential*, $Z_s$ is the effective nuclear charge and $\beta_{rs}$ is the *two-center resonance integral*, another parameter of the model.

@@

where $K$ is the exchange integral between the two orbitals involved in the single excitation (normally the HOMO and the LUMO). As defined here, a positive $\DeltaE{ST}$ corresponds to the usual situation with the triplet being lower in energy than the singlet, while a negative $\DeltaE{ST}$ corresponds to an inverted gap. The exchange integral depends strongly on the spatial overlap between the the two orbitals, which in the ZDO approximation can be computed as:[@peachExcitationEnergiesDensity2008]

@@

where $|c_{r,i}|$ is the absolute value of the coefficient of orbital $i$ on center $r$. When the overlap is zero, the exchange interaction vanishes and $\DeltaEST{SCF}$ is zero. Unfortunately, so is the oscillator strength, $f$, between the S~1~ excited state and the S~0~ ground state,[@klessingerExcitedStatesPhotochemistry1995] which is non-optimal for OLED materials that should emit light.[@gomez-bombarelliDesignEfficientMolecular2016] Consequently, there is a trade-off between having a small HOMO-LUMO overlap to reduce the exchange interaction, while still maintaining a sufficient oscillator strength.

@@

We used four separate datasets for this study. The first is a set of azaphenalenes previously studied by some of us, comprising 256 substituted compounds, for which excitation energies were computed with (2nd-order) algebraic-diagrammatic construction, {{< var reference_level >}}/cc-pVDZ, and geometries were optimized with B3LYP/cc-pVDZ[@polliceOrganicMoleculesInverted2021] The second is a set substituted 138 azaazulenes with {{< var reference_level >}}/cc-pVDZ excitation energies and optimized with the B97-3c composite method.[@nigamArtificialDesignOrganic2023] The third is a set of 16 rationally designed scaffolds, which including substitutions amounts to {{< var global_n_total >}} unique compounds, optimized at the B97-3c level and with excitation energies at the {{< var reference_level >}}/cc-pVDZ level.[@polliceRationalDesignOrganic2023] The fourth is a set 315 non-alternant hydrocarbons including substitutions, divided into three subsets of size 76, 187 and 52, optimized at the $\omega$B97X-D/def2-TZVP level and with excitation energies at the CC2/aug-cc-pVTZ level.

@@

## Orbital decomposition of the inverted gap

@@

@@

@@

Azaphenalene is the prototypical molecule that started the renewed investigations into INVEST molecules,[@desilvaInvertedSingletTriplet2019;@sobolewskiAreHeptazineBasedOrganic2021] and has been the subject of numerous studies with high-level quantum chemical methods. Can we capture the inverted gap, as first measured by Leupin and Wirz in 1980?[@leupinLowlyingElectronicallyExcited1980] The calculations reveal that the HOMO and LUMO are well separated with spatial overlap of {{< var azaphenalene_overlap >}} and a correspondingly low exchange interaction of only {{< var azaphenalene_2_exchange >}} kcal/mol ([@fig-azaphenalene-orbitals]). The larger azaphenalene has 30 excitations that could possibly contribute to the large $\DeltaE[DSP]{SCF}$ of {{< var azaphenalene_dsp_scf >}} kcal/mol, but the most important by far are those from the doubly degenerate HOMO-1 to the doubly degenerate LUMO+1 ([@fig-azaphenalene-excitations]). In total, $\DeltaEST[DSP]{SCF}=$ {{< var azaphenalene_t1_s1_dsp_scf >}} kcal/mol, which is slightly smaller at the CIS level with $\DeltaEST[DSP]{CIS}=$ {{< var azaphenalene_t1_s1_dsp_cis >}} kcal/mol.

To conclude this section on the model compounds, our PPP-based protocol is able to capture the inverted gap, and also provides qualitative understanding into the physical mechanism of dynamic spin polarization through identification and visualization of the corresponding excitations. For these two compounds, the calculated $\DeltaEST[DSP]{}$ are more negative at the SCF level compared to CIS. Compared to {{< var reference_level >}}, it would seem that CIS is preferable, but to elucidate which method is better, we now turn to larger datasets.

@@

While our method captures the inverted gap of azaphenalene and pentalene, it also needs to capture trends with substitution to work effectively in virtual screening. We therefore calculated 256 substituted azaphenalenes that have previously been studied by some of us.[@polliceOrganicMoleculesInverted2021] The PPP level S~1~ and T~1~ excitation energies are well correlated with those from {{< var reference_level >}} with $R^2$ of {{< var local_azaphenalenes_r2_singlet >}} and {{< var local_azaphenalenes_r2_triplet >}}, respectively ([@fig-local-azaphenalenes]a,b). Unfortunately, the oscillator strengths are not as well reproduced with an $R^2$ of {{< var local_azaphenalenes_r2_f >}}, although the Spearman $\rho$ of {{< var local_azaphenalenes_rho_f >}} indicates the values might be used to rank potential candidates ([@fig-local-azaphenalenes]c). The problem of obtaining accurate oscillator strengths at the PPP level of theory is well known in the literature, and is especially exacerbated with low oscillator strengths which are prevalent in this dataset.[@misurkinElectronicStructureConjugated1974] Crucially, the $\DeltaE{ST}$ shows good correlations, with $R^2=$ {{< var local_azaphenalenes_r2_cis >}} using $\DeltaEST[DSP]{CIS}$ ([@fig-local-azaphenalenes]f). This $R^2$ value is essentially unchanged from the gap at the CIS level without DSP ({{< var local_azaphenalenes_r2_cis_raw >}}, [@fig-local-azaphenalenes]d) and markedly better than for the gap at the SCF level ({{< var local_azaphenalenes_r2_scf >}}, [@fig-local-azaphenalenes]e), showing the importance of going beyond the SCF level to include at least some configuration interaction. The results are of similar or better quality recently achieved by Pernal and co-workers for a different set of azaphenalenes using SCF + DSP with orbitals from DFT.[@drwalRoleSpinPolarization2023] To further analyze the results in terms of a binary classification into normal/inverted, we calculated the true positives (TP), true negatives (TN), false positives (FP) and false negatives (FN), as well as a range of common classification scores (@tbl-local-azaphenalenes). For definitions, see [@eq-f1] to [@eq-roc-auc]. The most important metrics for virtual screening are, in our opinion, the recall and the specificity. The recall measures the fraction of inverted molecules that the protocol captures, and is {{< var local_azaphenalenes_recall_scf >}} at the SCF level and {{< var local_azaphenalenes_recall_cis >}} at the CIS level. The specificity measures the fraction of non-inverted molecules that the protocol identifies, and is {{< var local_azaphenalenes_specificity_scf >}} at the SCF level and {{< var local_azaphenalenes_specificity_cis >}} at the CIS level. These metrics mean that we are able to capture a large proportion of the molecules with inverted gaps, while filtering out most non-inverted molecules. Below we will show how we can improve the results even more with a linear correction to $\DeltaEST[DSP]{CIS}$ based on results from a wider set of compounds. The results for these azaphenalenes indicate that the PPP protocol could be used to prescreen candidates of this compound class for $\DeltaE{ST}$, while further pruning with another method is likely needed for the oscillator strengths.

![Excitation properties for azaphenalenes against the reference level. (a) Singlet excitation energies. (b) Triplet excitation energies. (c) Oscillator strengths. (d) Singlet-triplet energy gaps with CIS. (e) Singlet-triplet energy gaps with SCF + DSP. (f) Singlet-triplet energy gaps with CIS + DSP.]({{< var fig_local_azaphenalenes >}}.pdf){#fig-local-azaphenalenes}

@@

While we see success for the azaphenalenes, the situation for the azaazulenes is unfortunately worse. The PPP level S~1~ and T~1~ excitation energies are only moderately correlated with those from {{< var reference_level >}} with $R^2$ of {{< var local_azulenes_r2_singlet >}} and {{< var local_azulenes_r2_triplet >}}, respectively ([@fig-local-azulenes]a,b). For the oscillator strengths, the situation is rather catastrophic with $R^2$ of {{< var local_azulenes_r2_f >}} implying no correlation whatsoever ([@fig-local-azulenes]c). The situation for the $\DeltaE{ST}$ is also worse than for the azaphenalenes, with $R^2=$ {{< var local_azulenes_r2_cis >}} with CIS + DSP, a marked improvement over {{< var local_azulenes_r2_scf >}} with SCF + DSP ([@fig-local-azulenes]e,f). Unfortunately, the negative gaps are not recovered, leading to recalls of {{< var local_azulenes_recall_cis >}} as none of five inverted molecules could be identified ([@tbl-local-azulenes]). It could be speculated that the worse performance for the azaazulenes comes from the fact that they are non-alternants, for which the approximations are expected to be less applicable. Considering the contrasting performance for the azaphenalenes and azaazulenes, it seems clear that the PPP screening protocol will inevitably miss some hits belonging to particular compound classes. To investigate the global performance of the model we turned to a more diverse set with 16 different compound classes.

![Excitation properties for azaazulenes against the reference level. (a) Singlet excitation energies. (b) Triplet excitation energies. (c) Oscillator strengths. (d) Singlet-triplet energy gaps with CIS. (e) Singlet-triplet energy gaps with SCF + DSP. (f) Singlet-triplet energy gaps with CIS + DSP.]({{< var fig_local_azulenes >}}.pdf){#fig-local-azulenes}

@@

In this section, we used a dataset previously generated by some of us using rational design rules, comprising 16 different molecular scaffolds, the details of which will be presented elsewhere. Out of {{< var global_n_total >}} compounds, only {{< var global_n_failed >}} cyclobutadienes failed to compute as they had rearranged in the DFT simulations and our topology detection algorithm interpreted them as having formed transannular single bond. This corresponds to a success rate of {{< var global_percent_success >}}%. To put this into context, ~25% of the CASSCF calculations and 9% of the CIS calculations failed in a recent HTVS workflow by Troisi, Padula and co-workers.[@omarIdentificationUnknownInverted2023] The total runtime using four cores on the M2 processor was {{< var time_cpu_hours >}} CPU hours, with a mean time per compounds of {{< var time_mean >}} seconds. The S~1~ and T~1~ excitation energies are fairly reproduced with ([@fig-global-all]a,b) $R^2$ of {{< var global_all_r2_singlet >}} and {{< var global_all_r2_triplet >}} against {{< var reference_level >}}, respectively. As seen in [@fig-global-all-gaps]a, the $\DeltaE{ST}$ at the SCF + DSP level shows a moderate $R^2$ of {{< var global_all_r2_scf >}}, which increases significantly to {{< var global_all_r2_cis >}} at the CIS + DSP level ([@fig-global-all-gaps]b). However, due to a systematic overestimation of $\DeltaE{ST}$, the F1 score is still low at {{< var global_all_f1_cis >}} and the recall is only {{< var global_all_recall_cis >}} ([@tbl-global-all]). We therefore added a linear correction $\DeltaEST[DSP]{CIS,LC}=$ {{< var global_lc_coef >}} $\times\DeltaEST[DSP]{CIS}$ {{< var global_lc_intercept >}}, which increases the F1 score to {{< var global_all_f1_cis_corr >}} and the recall to {{< var global_all_recall_cis_corr >}} ([@fig-global-all-gaps]c). The specificity simultaneously decreases from {{< var global_all_specificity_cis >}} to {{< var global_all_specificity_cis_corr >}}, but overall the linear correction would be preferable for virtual screening where there is a strong focus on finding rare hits. Applying the linear correction to the azaphenalenes ([@fig-local-corr]a) and azaazulenes ([@fig-local-corr]b) leads to an improvement in the recall of the former from {{< var local_azaphenalenes_recall_cis >}} to {{< var local_azaphenalenes_recall_cis_corr >}} while still none of the inverted azaazulenes are recovered. The oscillator strengths are unfortunately poorly correlated for the rational design set, with $R^2=$ {{< var global_all_r2_oscillator >}}, although the Spearman rank correlation is more encouraging at {{< var global_all_rho_oscillator >}} ([@fig-global-all]c). Taken together, the results reinforce the conclusions from the case study of the azaphenalenes, that the PPP protocol is suitable for screening $\DeltaE{ST}$ while it struggles for oscillator strengths.
